---
title: "Working with Tables"
output: html_document
---

<!-- This file by Charlotte Wickham is licensed under a Creative Commons Attribution 4.0 International License, adapted from the orignal work at https://github.com/rstudio/master-the-tidyverse by RStudio. -->

```{r setup}
library(tidyverse)
library(haven)
library(readxl)
library(gapminder)
```

# Tabular Data

Let's look at just our tabular datasets again: 

```{r}
starwars
```

```{r}
gapminder
```



## Your turn 1

Take a look at these two datasets, by typing their names into the console:
* quakes
* mpg

## Your turn 2

Try running each line one by one, with Ctrl + Enter (or Cmd + Enter on Mac):

```{r}
dim(x = mpg)     
names(x = mpg)   
glimpse(x = mpg) 
View(x = mpg)  
```

## Your turn 3

How many rows of data are in quakes?

```{r}

```

What are the names of the variables in quakes?

```{r}

```

## Your turn 4

Run the chunk to get help on `mpg`. What is this data?
```{r}
?mpg
```

# Vector Data

## Your turn 5

Take another look at mpg.  What kind of data is in each column?

```{r}
mpg
```

# Importing Data

## Your Turn 6

Take a look in the data/ directory in the project folder.

Try reading in each of:

* `deaths.xls`
* `nimbus.csv`
* `iris.sav`

```{r}

```


## Your Turn 7

Can you see what is wrong with the Excel file when it is imported?

Scan the Arguments section of ?read_excel, can you find an argument that might help? Try it!

```{r}
?read_excel
```


<!-- This file by Charlotte Wickham is licensed under a Creative Commons Attribution 4.0 International License, adapted from the orignal work at https://github.com/rstudio/master-the-tidyverse by RStudio. -->

```{r setup}
library(tidyverse)
library(gapminder)

# Toy dataset to use
pollution <- tribble(
       ~city,   ~size, ~amount, 
  "New York", "large",      23,
  "New York", "small",      14,
    "London", "large",      22,
    "London", "small",      16,
   "Beijing", "large",      121,
   "Beijing", "small",      56
)
```

## gapminder

```{r}
gapminder
```

## Your Turn 1

* `filter()` selects rows
* logical tests

See if you can use the logical operators to manipulate our code below to show:

The data for Canada
```{r}
filter(gapminder, country == "New Zealand")
```

All data for countries in Oceania
```{r}
filter(gapminder, country == "New Zealand")
```

Rows where the life expectancy is greater than 82
```{r}
filter(gapminder, country == "New Zealand")
```


## Your Turn 2

Use Boolean operators to alter the code below to return only the rows that contain:

* Canada before 1970
```{r}
filter(gapminder, country == "New Zealand", year > 2000)
```

*  Countries where life expectancy in 2007 is below 50

```{r}
filter(gapminder, country == "New Zealand", year > 2000)
```

* Countries where life expectancy in 2007 is below 50, and are not in Africa.

```{r}
filter(gapminder, country == "New Zealand", year > 2000)
```

## Your Turn 3

Alter the code to:

* Add an africa column, which contains TRUE is the country is on the Africa continent.
* Add a rank_pop column to rank each row in gapminder from largest pop to smallest pop.


```{r}
mutate(gapminder)
```

## Your Turn 4

Use summarise() to compute three statistics about the data:

* The first (minimum) year in the dataset
* The last (maximum) year in the dataset
* The number of countries represented in the data (Hint: use cheatsheet)

```{r}
gapminder 
```

## Your Turn 5

Extract the rows where continent == "Africa" and year == 2007. 

Then use summarise() and summary functions to find:

1. The number of unique countries
2. The median life expectancy

```{r}
gapminder 
```


```{r}
gapminder
```
## Your Turn 6

Find the median life expectancy by continent

```{r}
gapminder 
```

## Your Turn 7

Brainstorm with your neighbor the sequence of operations to find:  the country with biggest jump in life expectancy  (between any two consecutive records) for each continent.
  
## Your Turn 8

Find the country with biggest jump in life expectancy (between any two consecutive records) for each continent.

```{r}


```

***

# Take aways

* Extract cases with `filter()`  
* Make new variables, with `mutate()`  
* Make tables of summaries with `summarise()`  
* Do groupwise operations with `group_by()`
* Connect operations with `%>%`  
